Estimating changes in temperature distributions in a large ensemble of climate simulations using quantile regression

### 2018-06-19 paper_revision.tex
Ryans and Michaels comments again

% XX "Study" was used in two different ways in first two sentences. I changed to say "we will talk about 3 things" - EJM
%One under-exploited advantage of ensembles of multiple simulations is that the increased volume of data allows %XX study of examining the entire distribution of a climate variable.
Large single model ensembles offer at least three advantages over a single simulation of a climate model. The most obvious advantage is that the increased data volume allows closer examination of the entire distribution of a climate variable.
 Studies of climate variability to date are generally divided between those that address the center of the distribution 
\citep[e.g.][]{semenov2002secular,raisanen2002co2, kitoh2009changes, screen2014arctic, schneider2015physics}, and those that address its tails \citep[e.g.][]{katz1992extreme, meehl2009relative, northrop2011threshold, davison2012statistical, huser2014space, trenberth2015attribution, huang2015estimating, jalbert2017spatiotemporal}, generally via extreme value theory. %In part the division has arisen from the data scarcity imposed by particular datasets.  %---- END CC
%High-resolution dynamically downscaled model studies \citep{XX,XX,XX} could in principle involve multiple runs, but their computational demands generally limit them to at most a few decades of present and future simulations. 
% XX Those studies 
A more limited body of studies address overall distributional changes in climate variables, but these generally focus on observations or observation-based data products, which are necessarily limited in terms of data amount and therefore require spatial or temporal aggregation \citep{donat2012shifting, stainforth2013mapping, chapman2013estimating, huybers2014us, mckinnon2016changing, rhines2017seasonally}. Aggregating data spatio-temporally requires stationarity assumptions of the signal or explicitly modeling the spatio-temporal structure. When studying model projections using ensembles, the large amounts of data at each location allows us to estimate changes in the distribution of climate variables (e.g. temperature) without spatial aggregation.

A second advantage provided by large single model ensembles is that trends in both means and variability need not be modeled as linear in time \citep{franzke2015local,gao2017quantile}. Forcings are not linear over centennial timescales, and a linear approximation can be misleading \citep[see for example][]{poppick2017estimating}. The increased data provided by ensembles allows more flexible statistical models to represent complex climate responses. %a realistic and often nonlinear scenario. 
As we will show, distributions of daily temperature evolve nonlinearly, and follow different trajectories for different quantiles (i.e. different parts of the distribution). %We therefore need a sound method that can
Analysis methods should therefore be able to take into account nonlinearities both in time and across quantiles.

@@

{While the focus on quantiles and quantile differences makes the use of quantile regression natural, it is possible to estimate quantiles using empirical distributions of temperatures within, say, seasonal blocks for each decade, and use these results to study how quantile differences change over time.  Even if ignoring changes in quantiles within a season is viewed as satisfactory, we prefer the quantile regression approach with carefully chosen covariates to capture seasonal and long-term trends outlined in Section 3.  For example, in Section 4, we show how the first day of the year at which a range of estimated quantiles cross certain temperature thresholds evolve over time at a site, which is not a question that can be answered when fitting a single temperature distribution for each season.}

@@

The ensemble is appropriate for the purpose of studying coupled internal climate variability because it is based on a ${\sim}10{,}000$ year pre-industrial control simulation. After a $\sim$4000 year spin-up using constant preindustrial conditions, 50 historical hindcasts (1850-2005) are initialized from snapshots of the coupled model state taken every 100 years, so that the last hindcast is initialized after approximately 9000 years of the control simulation. Each hindcast is then extended to 2100 using the Representative Concentration Pathway (RCP) 8.5 scenario. The 100-year gap between each new initialization ensures nearly independent ensemble members that fully capture internal variability within the coupled system. RCP8.5 corresponds to anthropogenic radiative forcing of roughly 8.5 W m$^{-2}$ by 2100 \citep{moss2010next}. More information about the model and ensemble design can be found in \citet{sriver2015effects}.

@@

% CC -- BEGIN edits below. "Validate" sounds a little more formal than what you're actually doing.
% CC -- if you say " With results for winter, each quantity shows a decent correspondence .." that implies that summer results DONT show a correspondence. 
%To evaluate whether the CESM simulations provide sufficiently realistic temperature distributions for the current analysis, 
{ In assessing the practical relevance of our results on the shapes of temperature distributions, it is perhaps helpful} to compare CESM temperatures with those from the ERA-Interim (European Reanalysis) data product \citep{dee2011era}. Figure \ref{model_reanalysis_world} shows the model/reanalysis comparison for winter; for summer see Supplementary Online Material Figure \ref{model_reanalysis_world_JJA}. The model underestimates variability in some places, and produces excessively cold winter temperatures in the Arctic. The resulting temperature gradients contribute to excess variability and negative skew in the northern mid-latitudes.
Skewness is proportional to the cube of temperature after subtracting off the average seasonal temperature; see Appendix \ref{app:moments}. 
{We see that, despite some discrepancies, Figure  \ref{model_reanalysis_world} shows the model runs do a good job overall of capturing where wintertime skewness is positive and negative.}
%% ejm changed from cubes to cube.
Throughout this work, we will show in-depth analysis from three locations with distinct temperature distributions to highlight our proposed method (\textbf{a}, \textbf{b}, and \textbf{c} shown in Figure \ref{model_reanalysis_world}). See Supplementary Online Material Figure \ref{rea_temp_locations} for comparison of model and reanalysis temperature distributions in both summer and winter for these locations. 
% CC could add a sentence saying this is still OK for your purpose if you want, but I think you can skip that

@@

To simplify notation, we construct a matrix $X$ where each column contains a basis function and each row refers to a unique value of $d, t$ and ensemble member. Using this matrix we construct our temperature model in vectorized form,
 \begin{equation}
 \label{eq:matrix_version}
 T_q = X\beta_q, 
 \end{equation}
 where $\beta_q$ contains the 32 basis coefficients $\alpha, a_i, b_j, c_{i,j}$. The predictor matrix $X$ has $32$ columns, each corresponding to one basis function, and $365 \times 250 \times 50$ rows. To get a confidence interval for each entry of $T_q$, we re-estimate the coefficients, $\beta_q$, using a resampled data set. Because we have 50 simulations we resample the data by drawing whole simulations from our ensemble of $50$ simulations. By resampling complete realizations, the dependency structure within realizations is maintained in the resampled data. Repeating this resampling and re-estimation procedure 100 times yields pointwise confidence bands around each estimated $T_q$. Appendix \ref{app:model}c provides further details about uncertainty quantification.

@@

To facilitate comparison with previous studies, we first perform a preliminary analysis where we replicate more standard methods. That is, we examine changes in the aggregate distribution of temperatures over multi-week and multi-month intervals, before we show results from our new approach that calculates responses for individual days. Even the standard analysis readily shows that temperature distributions in the CESM ensemble change over the RCP 8.5 scenario (Figures \ref{location_distributions}, \ref{meanStdSkew_Winter}, and \ref{meanStdSkew_Summer}, which compare the “initial” and “final” time windows 1850-1864 and 2086-2100). Means uniformly shift to warmer temperatures, but the shapes of the distributions also change in terms of variance and skewness. Figure \ref{location_distributions} shows initial and final distributions in our example locations for aggregated 15-day periods in winter and summer. In at least two of the three depicted locations, it is clear that the distributions are becoming narrower, although quantifying exactly how the tails are changing requires a quantification of the tail size and shape.
%difficult to quantify exactly how the tails are changing.

@@

%To what extent the distributional characteristics are changing, and whether these changes are significant. 
Our methodology for quantile estimation provides additional information that helps to
 quantify how temperature distributions are changing and to estimate the uncertainty associated with each change. 
%The goal is to understand how the temperature distribution is changing, both in the middle and in the extremes (or tails) of the distribution. 
We can evaluate not only bulk variability  -- the interquartile range (IQR), the difference between the $0.25$ and $0.75$ quantiles -- but differences between any two quantiles. 
{Denote by $\Delta q_{l}$ the difference between estimated 0.05 and 0.025 quantiles,  and $\Delta q_{h}$, the difference between estimated 0.975 and 0.95 quantiles.
These quantities measure tail variability in the same way that interquartile range measures the variability of the bulk distribution.
If the skewness of a distribution changes over time, then future distributions are not simply scaled versions of present distributions. That is, their tail variabilities must change by a different factor than the IQR.
  In the case of the northern mid-latitudes winter temperatures shown in Figure \ref{meanStdSkew_Winter}, where distributions become more negatively skewed as bulk variability decreases in the future, the effect could result from either a low tail contracting less than the bulk (or actually increasing), and/or a high tail contracting more than the bulk. Our methodology allows readily differentiating these cases.}

To assess whether the high tail and/or the low tail is driving changes in skewness, we consider the fractional changes in low, high, and bulk variability.
%Bulk variability, e.g. the interquartile range (IQR), is commonly linked to the tail variability. For example, a decrease in the bulk variability as a result of an overall scale change will produce a corresponding decrease in the tail variability. However, if for example the tail contracts less the bulk, there will be a skewness change. To show how variability at the different quantiles change with respect to the initial value, 
{Fixing day of the year, denote the initial and final quantile difference as $\Delta q_{r, i}$ and $\Delta q_{r, f}$, where $r=l,m,$ or $h$, indicated low tail, middle of the distribution (IQR) or high tail.  tThe temporal change in quantile differences relative to the initial year is then
\begin{equation}
	\rho = \frac{\Delta q_{r, f} - \Delta q_{r, i}}{\Delta q_{r, i}}.
\label{eq:relnorm}
\end{equation}
 Because we model the complete temperature distribution for each day of the year for all years, we choose representative days to understand winter and summer changes (Jan 1 and July 5, respectively), and consider the difference between the beginning and end of the scenarios, the years 1850 and 2000. For these representative days, we show in Figure \ref{winter_summer_var_avg} the fractional variability changes of $\rho$ for low and high tails as well as the IQR. }

@@

To clarify the relative contributions of high and low tails to skewness changes, we 
also examine evolving temperature variability in the bulk and tails as a function of seasonality as well as long term change. 
Figure \ref{dq_vs_day} shows absolute variability changes 
for the three example locations \textbf{a}, \textbf{b}, and \textbf{c}
estimated using our quantile model, and for fractional changes see Supplementary Online Material Figure \ref{dq_vs_day_norm}. The uncertainty around our estimates is quantified by resampling the original simulations (with replacement) and recomputing the estimates using this new set of simulations (see Appendix \ref{app:model}\ref{app:uq} for details). In all locations, wintertime skewness changes are driven by the relative changes in IQR and low tails. In the higher-latitude locations \textbf{a} and \textbf{b}, more negative winter skew results because the IQR contracts even more strongly than does the low tail variability. %In other words, the low tails ``stick''.  
In the lower-latitude location \textbf{c}, more positive winter skew results because the IQR changes slightly while the low tail variability contracts strongly.

@@

The complexity of the relationships in Figure \ref{dq_vs_day} also shows how misleading it may be to use a three-month block to represent a season. While all three locations show larger IQR in winter than summer, where the transition from winter to summer happens more quickly at some locations than at others. This transition takes place abruptly in the northernmost location \textbf{a} and more gradually in \textbf{c}. Low-tail variability seasonal transitions are even sharper than those of IQR in \textbf{a} and \textbf{b}, but more gradual in \textbf{c}.
%also has very different profiles between each location. Location \textbf{a} hits peak low-tail variability during the shoulder seasons (spring and fall), while location \textbf{b} shows a more gradual low tail variability rise toward winter. Location \textbf{c} shows less temporal change in the IQR compared with the other locations. Overall, 
In contrast, high-tail variability is more seasonally constant overall than low-tail variability. 
Through these examples, we see how our method offers detailed information about changes in variability across seasons and annual change, usually unavailable when analyzing each season separately.

While we show only three locations in the text here, an online interactive application allows similar in-depth examination of changes in model temperature distributions at all locations within North America, available at {\small\texttt{https://matzhaugen.com/links.html}}. The application allows the user to browse through any desired location to see how the variability changes as a function of season, year and quantile difference. We include temperature histograms of the first and last simulation year for the designated location, as well as maps that show the variability change spatially.

@@

The techniques presented in this study replicate several prior conclusions made in the literature, e.g. the well-known projected decrease in winter variability in the northern mid-latitudes \citep[e.g.][]{schneider2015physics} most likely due to amplified warming in the arctic \citep{screen2014arctic}. {We use differences between two quantiles in the same tail of the distribution as a tool for studying more nuanced changes in distributions. Quantile regression makes it possible to give accurate estimates of these quantiles that differ for every day and every year under study.  Resampling ensemble members yields confidence intervals for all quantities we estimate without having to model temporal dependence within model runs.} In the case study of CESM runs analyzed here, we relate the changes in tail variability to changes in skewness of the temperature distributions, and we find that in most of the domain analyzed, wintertime skewness changes are driven largely by the relative behavior of IQR and low tails. For example, in much of the continental northern U.S.\ and Canada, the low tail of temperature contracts substantially less than does the overall temperature variability.

These results may inform physical explanations for 
%The CESM study also usefully highlights other interesting changes, including 
the projection that skewness in winter temperature changes in a dipole pattern across North America. It is possible that the skewness change is a result of a change in the mean location and variability of the mid-latitude jet stream \citep[e.g.][]{barnes2013response}; this possibility may warrant further study.

The abundance of data available in large single model ensembles relative to single simulations allows using quantile regression to accurately estimate high quantiles within a single model structure, avoiding some of the limitations of extreme value theory. Unlike quantile regression, methods using extreme value theory require making assumptions about the shape of the tail of the distribution. {Of course, if one were interested in very extreme quantiles, say, below 0.001 or above 0.999, quantile regression estimates will become very noisy even with a 50-member ensemble, and approaches using extreme value theory may become necessary.}

By parameterizing the seasonally time-varying distribution of temperature through smooth functions using the whole year as our domain, we reveal previously unavailable details  about seasonal transitions. For example, we show here that springtime variability decreases occur later in the year at lower latitudes, and that seasonal transitions in tail variability differ from those in IQR.  While we analyze only temperature here, our method is intended to be general enough to be applied to other climate variables such as precipitation or humidity.
These detailed insights into climate variable distributions may be valuable for risk assessment studies that emphasize extreme events.

>>>>

% XX "Study" was used in two different ways in first two sentences. I changed to say "we will talk about 3 things" - EJM
%One under-exploited advantage of ensembles of multiple simulations is that the increased volume of data allows %XX study of examining the entire distribution of a climate variable.
Large single model ensembles offer at least three advantages over a single simulation of a climate model. The most obvious advantage is that the increased data volume allows closer examination of the entire distribution of a climate variable.
 Studies of climate variability to date address both the center of the distribution 
\citep[e.g.][]{semenov2002secular,raisanen2002co2, kitoh2009changes, screen2014arctic, schneider2015physics} and its tails \citep[e.g.][]{katz1992extreme, meehl2009relative, northrop2011threshold, davison2012statistical, huser2014space, trenberth2015attribution, huang2015estimating, jalbert2017spatiotemporal}, where the latter is often studied via extreme value theory. \citet{fischer2009future} use an ensemble to study seasonal variability in a way that accounts for intraseasonal changes in mean temperature.
A more limited body of studies address overall distributional changes in climate variables, but these generally focus on observations or observation-based data products, which are necessarily limited in terms of data amount and therefore require spatial or temporal aggregation \citep{donat2012shifting, stainforth2013mapping, chapman2013estimating, huybers2014us, mckinnon2016changing, rhines2017seasonally}. Aggregating data spatio-temporally requires stationarity assumptions of the signal or explicitly modeling the spatio-temporal structure. When studying model projections using ensembles, the large amounts of data at each location allows us to estimate changes in the distribution of climate variables (e.g. temperature) without spatial aggregation. For example, \citet{ylhaisi2014twenty} use CMIP3 data to remove a trend and a seasonal cycle and then take differences of quantiles of the residual temperature to measure variability and skewness. Unlike the work here, the seasonal cycle is not allowed to evolve over time and the quantiles of the residual temperautre are assumed constant within seasons and over 20 year blocks.

A second advantage provided by large single model ensembles is that trends in both means and variability need not be modeled as linear in time \citep{franzke2015local,gao2017quantile}. Forcings are not linear over centennial timescales, and a linear approximation can be misleading \citep[see for example][]{poppick2017estimating}. The increased data provided by ensembles allows more flexible statistical models to represent complex climate responses. 
As we will show, distributions of daily temperature evolve nonlinearly, and follow different trajectories for different quantiles (i.e. different parts of the distribution). 
Analysis methods should therefore be able to take into account nonlinearities both in time and across quantiles.

@@

{While the focus on quantiles and quantile differences makes the use of quantile regression natural, it is possible to estimate quantiles using empirical distributions of temperatures within, say, seasonal blocks for each decade, and use these results to study how quantile differences change over time.  Even if ignoring changes in quantiles within a season is viewed as satisfactory, we prefer the quantile regression approach with carefully chosen covariates to capture seasonal and long-term trends outlined in Section 3.  For example, in Section 4, we show how the first day of the year at which estimated quantiles cross certain temperature thresholds evolves over time at a site, which is not a question that can be answered when fitting a single temperature distribution for each season.}

@@

The ensemble is appropriate for the purpose of studying coupled internal climate variability because it is based on a ${\sim}10{,}000$ year pre-industrial control simulation. After a $\sim$4000 year spin-up using constant preindustrial conditions, 50 historical hindcasts (1850-2005) are initialized from snapshots of the coupled model state taken every 100 years, so that the last hindcast is initialized after approximately 9000 years of the control simulation. Each hindcast is then extended to 2100 using the Representative Concentration Pathway (RCP) 8.5 scenario. The 100-year gap between each new initialization ensures nearly independent ensemble members that fully capture internal variability within the coupled system. RCP8.5 corresponds to anthropogenic radiative forcing of roughly 8.5 W m$^{-2}$ by 2100 \citep{moss2010next}. More information about the model and ensemble design can be found in \citet{sriver2015effects, hogan2017analyzing, vega2017analysis}.

@@

% CC -- BEGIN edits below. "Validate" sounds a little more formal than what you're actually doing.
% CC -- if you say " With results for winter, each quantity shows a decent correspondence .." that implies that summer results DONT show a correspondence. 
%To evaluate whether the CESM simulations provide sufficiently realistic temperature distributions for the current analysis, 
{ In assessing the practical relevance of our results on the shapes of temperature distributions, it is perhaps helpful} to compare CESM temperatures with those from the ERA-Interim (European Reanalysis) data product \citep{dee2011era}. Figure \ref{model_reanalysis_world} shows the model/reanalysis comparison for winter; for summer see Supplementary Online Material Figure \ref{model_reanalysis_world_JJA}. The model underestimates variability in some places, and produces excessively cold winter temperatures in the Arctic. The resulting temperature gradients contribute to excess variability and negative skew in the northern mid-latitudes.
Skewness is proportional to the cube of temperature after subtracting off the average seasonal temperature; see Appendix \ref{app:moments}. 
{We see that, despite some discrepancies, Figure \ref{model_reanalysis_world} shows that the model exhibits skill in capturing where wintertime skewness is positive and negative.}
%% ejm changed from cubes to cube.
Throughout this work, we will show in-depth analysis from three locations with distinct temperature distributions to highlight our proposed method (\textbf{a}, \textbf{b}, and \textbf{c} shown in Figure \ref{model_reanalysis_world}). See Supplementary Online Material Figure \ref{rea_temp_locations} for comparison of model and reanalysis temperature distributions in both summer and winter for these locations. 
% CC could add a sentence saying this is still OK for your purpose if you want, but I think you can skip that

@@

To simplify notation, consider a matrix $X$ where each column contains a basis function and each row refers to a unique value of $d, t$ and ensemble member. Using this matrix we construct our temperature model in vectorized form,
 \begin{equation}
 \label{eq:matrix_version}
 T_q = X\beta_q, 
 \end{equation}
 where $\beta_q$ contains the 32 basis coefficients $\alpha, a_i, b_j, c_{i,j}$. The predictor matrix $X$ has $32$ columns, each corresponding to one basis function, and $365 \times 250 \times 50$ rows. To get a confidence interval for each entry of $T_q$, we re-estimate the coefficients, $\beta_q$, using a resampled data set. Because we have 50 simulations we resample the data by drawing whole simulations from our ensemble of $50$ simulations. By resampling complete realizations, the dependency structure within realizations is maintained in the resampled data. Repeating this resampling and re-estimation procedure 100 times yields pointwise confidence bands around each estimated $T_q$. Appendix \ref{app:model}c provides further details about uncertainty quantification.

@@

To facilitate comparison with previous studies, we first perform a preliminary analysis where we replicate more standard methods. That is, we examine changes in the aggregate distribution of temperatures over multi-week and multi-month intervals, before we show results from our new approach that calculates responses for individual days. The standard analysis readily shows that temperature distributions in the CESM ensemble change over the RCP 8.5 scenario (Figures \ref{location_distributions}, \ref{meanStdSkew_Winter}, and \ref{meanStdSkew_Summer}, which compare the “initial” and “final” time windows 1850-1864 and 2086-2100). Means uniformly shift to warmer temperatures, but the shapes of the distributions also change in terms of variance and skewness. Figure \ref{location_distributions} shows initial and final distributions in our example locations for aggregated 15-day periods in winter and summer. In at least two of the three depicted locations, it is clear that the distributions are becoming narrower, although quantifying exactly how the tails are changing requires a quantification of the tail size and shape.
%difficult to quantify exactly how the tails are changing.

@@

%To what extent the distributional characteristics are changing, and whether these changes are significant. 
Our methodology for quantile estimation provides additional information that helps to
 quantify how temperature distributions are changing and to estimate the uncertainty associated with each change. 
%The goal is to understand how the temperature distribution is changing, both in the middle and in the extremes (or tails) of the distribution. 
We can evaluate not only bulk variability  -- the interquartile range (IQR), the difference between the $0.25$ and $0.75$ quantiles -- but differences between any two quantiles. 
{Denote by $\Delta q_{l}$ the difference between estimated 0.05 and 0.025 quantiles,  and $\Delta q_{h}$, the difference between estimated 0.975 and 0.95 quantiles.
These quantities measure tail variability in the same way that interquartile range measures the variability of the bulk distribution.
If the skewness of a distribution changes over time, then future distributions are not simply scaled versions of present distributions. That is, their tail variabilities must change by a different factor than the IQR.
  In the case of the northern mid-latitudes winter temperatures shown in Figure \ref{meanStdSkew_Winter}, where distributions become more negatively skewed as bulk variability decreases in the future, the effect could result from either a low tail contracting less than the bulk (or actually increasing), and/or a high tail contracting more than the bulk. Our methodology allows for differentiating these cases.}

To assess whether the high tail and/or the low tail is driving changes in skewness, we consider the fractional changes in low, high, and bulk variability.
%Bulk variability, e.g. the interquartile range (IQR), is commonly linked to the tail variability. For example, a decrease in the bulk variability as a result of an overall scale change will produce a corresponding decrease in the tail variability. However, if for example the tail contracts less the bulk, there will be a skewness change. To show how variability at the different quantiles change with respect to the initial value, 
{Fixing day of the year, denote the initial and final quantile difference as $\Delta q_{r, i}$ and $\Delta q_{r, f}$, where $r=l,m,$ or $h$, indicated low tail, middle of the distribution (IQR) or high tail. The temporal change in quantile differences relative to the initial year is then
\begin{equation}
	\rho = \frac{\Delta q_{r, f} - \Delta q_{r, i}}{\Delta q_{r, i}}.
\label{eq:relnorm}
\end{equation}
 Because we model the complete temperature distribution for each day of the year for all years, we choose representative days to understand winter and summer changes (Jan 1 and July 5, respectively). Thus, we consider the difference between the beginning and end of the long-term climate scenarios, i.e. the years 1850 and 2100. For the representative days, we show in Figure \ref{winter_summer_var_avg} the fractional variability changes of $\rho$ for low and high tails as well as the IQR. }

@@

To clarify the relative contributions of high and low tails to skewness changes, we 
also examine evolving temperature variability in the bulk and tails as a function of seasonality as well as long term change. 
Figure \ref{dq_vs_day} shows absolute variability changes 
for the three example locations \textbf{a}, \textbf{b}, and \textbf{c}
estimated using our quantile model; for fractional changes see Supplementary Online Material Figure \ref{dq_vs_day_norm}. The uncertainty around our estimates is quantified by resampling the original simulations (with replacement) and recomputing the estimates using this new set of simulations (see Appendix \ref{app:model}\ref{app:uq} for details). In all locations, wintertime skewness changes are driven by the relative changes in IQR and low tails. In the higher-latitude locations \textbf{a} and \textbf{b}, more negative winter skew results because the IQR contracts even more strongly than does the low tail variability. %In other words, the low tails ``stick''.  
In the lower-latitude location \textbf{c}, more positive winter skew results because the IQR changes slightly while the low tail variability contracts strongly.

@@

The complexity of the relationships in Figure \ref{dq_vs_day} also shows how misleading it may be to use a three-month block to represent a season. While all three locations show larger IQR in winter than summer, the transition from winter to summer happens more quickly at some locations than at others. This transition takes place abruptly in the northernmost location \textbf{a} and more gradually in \textbf{c}. Low-tail variability seasonal transitions are even sharper than those of IQR in \textbf{a} and \textbf{b}, but more gradual in \textbf{c}.
%also has very different profiles between each location. Location \textbf{a} hits peak low-tail variability during the shoulder seasons (spring and fall), while location \textbf{b} shows a more gradual low tail variability rise toward winter. Location \textbf{c} shows less temporal change in the IQR compared with the other locations. Overall, 
In contrast, high-tail variability is more seasonally constant overall than low-tail variability.

While we show only three locations in the text here, an online interactive application allows similar in-depth examination of changes in model temperature distributions at all locations within North America, available at {\small\texttt{http://statvm1.cs.uchicago.edu/mahaugen/extremes/}}. The application allows the user to browse through any desired location to see how the variability changes as a function of season, year and quantile difference. We include temperature histograms of the first and last simulation year for the designated location, as well as maps that show the variability change spatially.

@@

The techniques presented in this study replicate several prior conclusions made in the literature, e.g. the well-known projected decrease in winter variability in the northern mid-latitudes \citep[e.g.][]{schneider2015physics} most likely due to amplified warming in the arctic \citep{screen2014arctic}. {We use differences between two quantiles in the same tail of the distribution as a tool for studying more nuanced changes in distributions. Resampling ensemble members yields confidence intervals for all quantities we estimate without having to model temporal dependence within model runs.} In the case study of CESM runs analyzed here, we relate the changes in tail variability to changes in skewness of the temperature distributions, and we find that in most of the domain analyzed, wintertime skewness changes are driven largely by the relative behavior of IQR and low tails. For example, in much of the continental northern U.S.\ and Canada, the low tail of temperature contracts substantially less than does the overall temperature variability.

These results may inform physical explanations for 
%The CESM study also usefully highlights other interesting changes, including 
the projection that skewness in winter temperature changes in a dipole pattern across North America. It is possible that the skewness change is a result of a change in the mean location and variability of the mid-latitude jet stream \citep[e.g.][]{barnes2013response}; this possibility may warrant further study. Furthermore, our approach allows us to examine the effect of polar amplification on tails (e.g. skewness) more comprehensively than usual methods using multi-model ensembles or models with fewer simulations (requiring more statistical assumptions).

The abundance of data available in large single model ensembles relative to single simulations also allows using quantile regression to estimate high quantiles accurately within a single model structure, avoiding some of the limitations of extreme value theory. Unlike quantile regression, methods using extreme value theory require making assumptions about the shape of the tail of the distribution. {Of course, if one were interested in very extreme quantiles, say, below 0.001 or above 0.999, quantile regression estimates could become very noisy even with a 50-member ensemble, and approaches using extreme value theory may become necessary.}

By parameterizing the seasonally time-varying distribution of temperature through smooth functions using the whole year as our domain, we reveal previously unavailable details  about seasonal transitions. For example, we show here that springtime variability decreases occur later in the year at lower latitudes, and that seasonal transitions in tail variability differ from those in IQR. There are other methods besides quantile regression one could use to estimate quantiles on a particular day $d$ and year $t$, for example, using empirical quantiles for all days of the year within a week of $d$ and all years within 5 years of $t$. Such possibilities are certainly worth exploring, although we prefer the flexibility of our approach in which we allow different numbers of basis functions for seasonality, long-term trends, and changes in seasonality. While we analyze only temperature here, our method is intended to be general enough to be applied to other climate variables such as precipitation or humidity.
These detailed insights into climate variable distributions may be valuable for risk assessment studies that emphasize extreme events.
